Choosing Color Palettes for Statistical Graphics

### 2006-12-30 trunk/vcd/paper/Color/TAS.Rnw
started TAS revision: non-technical with bad/good examples
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/vcd@634 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

\documentclass{Z}
%% need no \usepackage{Sweave}
\SweaveOpts{engine=R, eps=FALSE, echo=FALSE, results=hide}

@@

<<preliminaries>>=
library("vcd")
library("e1071")
library("mvtnorm")
library("ellipse")
library("kernlab")
library("KernSmooth")

source("functions.R")

@@

Color is an integral element of graphical displays in general, and many statistical
graphics in particular. Statistical software packages typically provide various color
palettes and allow practitioners to employ these for visualizing data in various types
of displays. However, more often than not, there is relatively little guidance about
how to choose an appropriate palette for a certain visualization task. Here, we address
this problem by suggesting a color selection strategy for visualizing both categorical
and numerical information by selecting colors along axes in a color space whose axes
can be matched with perceptual axes of the human visual system.

@@

Despite these known drawbacks, HSV space is still the predominantly-used color model in
color picker tools \citep{vcd:Moretti+Lyons:2002,vcd:Meier+Spalter+Karelitz:2004}
and the basis for many palettes in statistical software packages. A notable
exception is \pkg{ColorBrewer.org} \citep{vcd:Harrower+Brewer:2003},
an online tool for selecting color schemes for maps. It provides a rich
collection of prefabricated palettes (with a fixed maximal number
of colors) and guides its users how to choose a suitable palette for coding
various types of information. Here, we take a similar approach
and describe strategies for the choice of color palettes for categorical and numerical
data. Following \cite{vcd:Brewer:1999} and \cite{vcd:Harrower+Brewer:2003}, we distinguish
three types of palettes: qualitative, sequential and diverging. The first is tailored
for coding categorical information and the latter two are aimed at numerical variables
(or ordinal variables if some numerical coding for its levels is used).
Unlike \pkg{ColorBrewer.org}, we do not only provide fixed sets of colors but suggest
a general principle for selecting colors by traversing paths along perceptual axes in
a suitable color space. Consequently, the user can decide which path exactly should be
taken and how many colors should be selected. The color space employed by us is the
Hue-Chroma-Luminance (HCL) color space \citep[see][]{vcd:Ihaka:2003} and we show how
the general principles suggested can be turned into formulas describing paths of
coordinates in this space. Given a mapping from HCL to RGB coordinates, our formulas are
extremely easy to implement in software and we provide such an implementation
in the \proglang{R} language \citep{vcd:R:2006} using the powerful \proglang{R} graphics
system \citep[see][]{vcd:Murrell:2006} and the HCL
color implementation from package \pkg{colorspace} \citep{vcd:Ihaka:2004}. This gives the user
both a conceptual and computational tool box for experimenting with color palettes for
a particular display.

The remainder of the paper is organized as follows: Section~\ref{sec:spaces} provides
further information on HCL and other color spaces, contrasting in particular the properties
of HSV and HCL colors. Section~\ref{sec:palettes} introduces HCL-based qualitative,
sequential, and diverging palettes and Section~\ref{sec:illustrations} shows some brief
illustrations for each type of palette in statistical graphics. Section~\ref{sec:software}
provides some general remarks on the implementation in statistical software packages
as well as some details on our implementation in \proglang{R}. Section~\ref{sec:discussion}
concludes the paper with a discussion.

@@

\setkeys{Gin}{width=0.8\textwidth}
\begin{figure}[p]
\begin{center}
<<hsv-hcl>>=
if(generate_png) {
  files <- c("hcl-dcone", "hsv-cylinder")
  PNG <- function(x) paste(x, ".png", sep = "")

  png(file = PNG(files[1]), height = 400, width = 600)
  hcl.wheel(luminance = seq(5, 95, length = 6))
  dev.off()
  png(file = PNG(files[2]), height = 400, width = 600)
  hsv.wheel(value = seq(0, 1, length = 6), cone = TRUE)
  dev.off()
}
@
\includegraphics{hsv-cylinder}

\vspace*{2.5cm}

\includegraphics{hcl-dcone}
\caption{\label{fig:hsv-hcl} HSV and HCL space. The circles show the hue/saturation
and hue/chroma plane, respectively, for varying levels of value and luminance.}
\end{center}
\end{figure}

\begin{figure}[p]
\begin{center}
<<hsv-hcl-gray>>=
if(generate_png) {
  for(i in files) {
    file.copy(PNG(i), PNG(paste(i, "-gray", sep = "")), overwrite = TRUE)
    system(paste("mogrify -type Grayscale", PNG(paste(i, "-gray", sep = ""))))
  }
}
@
\includegraphics{hsv-cylinder-gray}

\vspace*{2.5cm}

\includegraphics{hcl-dcone-gray}
\caption{\label{fig:hsv-hcl-gray} HSV and HCL space (in gray levels). The circles
show the hue/saturation and hue/chroma plane, respectively, for varying levels of
value and luminance.}
\end{center}
\end{figure}

@@

Due to these three color axes, colors are typically described as
locations in 3-dimensional spaces. However, human perception of color
does not correspond to the physiological axes above, but rather to polar
coordinates in the color plane (yellow/blue vs.~green/red) plus a third
light/dark axis. Thus, perceptually-based color spaces try to capture the
perceptual axes
\begin{enumerate}
  \item \emph{hue} (dominant wavelength)
  \item \emph{chroma} (colorfulness, intensity of color as compared to gray)
  \item \emph{luminance} (brightness, amount of gray)
\end{enumerate}

A popular implementation of such a color space, available in many graphics and
statistics software packages, are HSV colors. They are a simple transformation of RGB colors
and are defined by a triplet $(H, S, V)$ with $H \in [0, 360]$ and $S, V \in [0, 100]$.
HSV space has the shape of a single regular cone (often inflated to a regular
cylinder). Vertical sections through this space are shown in the upper panel
of Figure~\ref{fig:hsv-hcl}, depicting hue and saturation given different value levels.
Although simple to specify and easily available in many computing environments,
HSV colors have a fundamental drawback: its three dimensions map to the three
dimensions of human color perception very poorly. The three dimensions are
confounded which is most easily seen when converting the vertical sections to 
gray scale images in Figure~\ref{fig:hsv-hcl-gray}. Clearly, the brightness of
colors is not uniform over hues and saturations (given value)---therefore,
HSV colors are often not considered to be perceptually based.

To overcome these drawbacks, various color spaces have been suggested
that properly map to the perception dimensions, the most prominent of which are
the CIELUV and CIELAB spaces developed by the \cite{vcd:CIE:2004}. \cite{vcd:Ihaka:2003}
argues that CIELUV colors are typically preferred for use with emissive technologies
such as computer screens which makes them an obvious candidate for implementation
in statistical software packages. By taking polar coordinates in the UV plane of
CIELUV, HCL colors are obtained, defined by
a triplet $(H, C, L)$ with $H \in [0, 360]$ and $C, L \in [0, 100]$. HCL space
has the shape of a distorted double cone: the admissible combinations of
chroma and luminance coordinates (within the space's boundaries) depend on the hue chosen.
The lower panel of Figure~\ref{fig:hsv-hcl} shows
vertical sections through this space: each of the resulting hue/chroma planes (given
luminance) is now properly balanced towards the same gray (going from black to white
with increasing luminance) which becomes obvious when converting the colors to
a gray scale as in Figure~\ref{fig:hsv-hcl-gray}. This balancing of HCL colors
gives us the opportunity to conveniently choose color palettes which code categorical
and/or numerical information by translating it to paths along the three perceptual axes.
However, some care is required for dealing with the irregular shape of the HCL space
which will be addressed in the following sections.

@@

@@

\setkeys{Gin}{width=0.8\textwidth}
\begin{figure}[p]
\begin{center}
<<pal-q-examples, fig=TRUE, height=4, width=6>>=
par(mfrow = c(2, 2), mar = c(0, 0, 3, 0))
pal(rainbow_hcl(4, start = 30, end = 300), main = "dynamic [30, 300]")
pal(rainbow_hcl(4, start = 60, end = 240), main = "harmonic [60, 240]")
pal(rainbow_hcl(4, start = 270, end = 150), main = "cold [270, 150]")
pal(rainbow_hcl(4, start = 90, end = -30), main = "warm [90, -30]")
@
\caption{\label{fig:pal-q-examples} Examples for qualitative palettes. Hue is varied
in different intervals for given $C = 50$ and $L = 70$.}
\end{center}
\end{figure}

@@

\setkeys{Gin}{width=.8\textwidth}
\begin{figure}[p]
\begin{center}
<<pal-s-examples, fig=TRUE, height=5, width=6>>=
par(mfrow = c(5, 1), mar = c(0, 0, 0, 0))
pal(sequential_hcl(12, c = 0, power = 2.2))
pal(sequential_hcl(12, power = 2.2))
pal(heat_hcl(12, c = c(80, 30), l = c(30, 90), power = c(1/5, 2)))
pal(terrain_hcl(12, c = c(65, 0), l = c(45, 90), power = c(1/2, 1.5)))
pal(rev(heat_hcl(12, h = c(0, -100), l = c(75, 40), c = c(40, 80), power = 1)))
@
\caption{\label{fig:pal-s-examples} Examples for sequential palettes, varying
only luminance (first panel), chroma and luminance (second panel), and hue,
chroma and luminance (remaining panels).}
\end{center}
\end{figure}

@@

\section{Illustrations} \label{sec:illustrations}

>>>>

\documentclass{Z}
%% need no \usepackage{Sweave}
\SweaveOpts{engine=R, eps=FALSE, echo=FALSE, results=hide}
\usepackage{amsfonts}

@@

<<preliminaries>>=
library("vcd")
library("mvtnorm")
library("KernSmooth")

source("functions.R")
source("Nigeria/bnd.R")

@@

Color is an integral element of graphical displays in general, and many statistical
graphics in particular. Statistical software packages typically provide various color
palettes and allow practitioners to employ these for visualizing data in various types
of displays. However, more often than not, there is relatively little guidance about
how to choose a suitable palette for a certain visualization task. Consequently,
sub-optimal color palettes are used for too many color graphics displayed in journal articles
or on presentation slides which could easily be enhanced if the underlying information
were captured appropriately in the color schemes. Here, we address
this problem by suggesting a color selection strategy for visualizing both categorical
and numerical information by selecting colors along axes in a color space whose axes
can be matched with perceptual axes of the human visual system.

@@

Despite these known drawbacks, HSV space is still the predominantly-used color model in
color picker tools \citep{vcd:Moretti+Lyons:2002,vcd:Meier+Spalter+Karelitz:2004}
and the basis for many palettes in statistical software packages. A notable
exception is \pkg{ColorBrewer.org} \citep{vcd:Harrower+Brewer:2003},
an online tool for selecting color schemes for maps. It provides a rich
collection of prefabricated palettes (with a fixed maximal number
of colors) and guides its users how to choose a suitable palette for coding
various types of information. Here, we take a similar approach
and describe strategies for the choice of color palettes for categorical and numerical
data. Following \cite{vcd:Brewer:1999} and \cite{vcd:Harrower+Brewer:2003}, we distinguish
three types of palettes: qualitative, sequential and diverging. The first is tailored
for coding categorical information and the latter two are aimed at numerical variables
(or ordinal variables if some numerical coding for its levels is used).
Unlike \pkg{ColorBrewer.org}, we do not only provide fixed sets of colors but suggest
a general principle for selecting colors by traversing paths along perceptual axes in
a suitable color space. Consequently, the user can decide which path exactly should be
taken and how many colors should be selected. The color space employed by us is the
Hue-Chroma-Luminance (HCL) color space \citep[see][]{vcd:Ihaka:2003} and we show how
the general principles suggested can be turned into formulas describing paths of
coordinates in this space. Given a mapping from HCL to RGB coordinates, our formulas are
extremely easy to implement in software and we provide such an implementation
in the \proglang{R} language \citep{vcd:R:2006} using the powerful \proglang{R} graphics
system \citep[see][]{vcd:Murrell:2006} and the HCL
color implementation from package \pkg{colorspace} \citep{vcd:Ihaka:2006}. This gives the user
both a conceptual and computational tool box for experimenting with color palettes for
a particular display.

The remainder of the paper is organized as follows: Section~\ref{sec:motivation} provides
several motivating examples, showing how typical HSV-based graphics can be enhanced
by using HCL-based palettes. Section~\ref{sec:spaces} gives a brief introduction
to the underlying HSV and HCL color spaces before Section~\ref{sec:palettes} 
suggests strategies for deriving HCL-based qualitative, sequential, and diverging palettes.
Section~\ref{sec:software} offers some general remarks on the implementation in statistical
software packages as well as some details on our implementation in \proglang{R}.
Section~\ref{sec:discussion} concludes the paper with a discussion.

\section{Motivation} \label{sec:motivation}

To show what can be gained by selecting appropriate color schemes, we present a collection
of illustrations with typical color graphics based on HSV palettes (always in the left panels)
and more suitable HCL palettes (always in the right panels). All HSV color palettes (and some
of the data sets) are taken from recent publications in statistical journals such as
\textit{Journal of the American Statistical Association},
\textit{Journal of Computational and Graphical Statistics} or
\textit{Computational Statistics \& Data Analysis}. The examples have been selected to provide
an overview of (typically colored) statistical graphics and their respective pitfalls (when
it comes to color palette choice). All graphics are in principle very useful but can be even enhanced
by using better color schemes.

Our first illustration is a heatmap, a very popular display for visualizing a function of
two arguments. Here, we use a bivariate kernel density estimate \citep[for the Old
Faithful geyser eruptions data from][]{vcd:Azzalini+Bowman:1990}---other typical applications
include objective functions with two arguments \citep[as in][]{vcd:Gneiting+Sevcikova+Percival:2006}
or (physical) measurements on a 2-dimensional grid \citep[as in][]{vcd:Yang+Buckley+Dudoit:2002}.
Figure~\ref{fig:bkde} shows such a heatmap bringing out the relationship between the duration
of an eruption of the Old Faithful geyser in Yellowstone National Park 
and the waiting time for this eruption. It reveals a multi-modal distribution: short waiting
times (around 50 minutes) are typically followed by a long eruption (around 4 minutes)
whereas long waiting times (around 80 minutes) can be followed by either a long
or short eruption (around 4 minutes).

\setkeys{Gin}{width=\textwidth}
\begin{figure}[t!]
\begin{center}
<<bkde>>=
if(generate_png) {
  data("geyser", package = "MASS")
  ## geyser2 <- cbind(geyser$duration[-299], geyser$waiting[-1])
  geyser2 <- cbind(geyser$duration, geyser$waiting)
  dens <- bkde2D(geyser2, bandwidth = c(0.2, 3), gridsize = c(201, 201))
  plot_bkde2D <- function(x, col = NULL, xlab = "duration", ylab = "waiting time", legend = FALSE, ...) {
    if(is.null(col)) col <- rev(grey.colors(33))
    if(legend) {
      filled.contour(x$x1, x$x2, x$fhat, col = col, nlevels = length(col), xlab = xlab, ylab = ylab, ...)
    } else {
      image(x$x1, x$x2, x$fhat,
        breaks = seq(from = min(x$fhat), to = max(x$fhat), length.out = length(col) + 1),
        xlab = xlab, ylab = ylab, col = col, axes = FALSE, cex.lab = 1.5, ...)
      box()
      axis(1, cex.axis = 1.5)
      axis(2, cex.axis = 1.5)
    }
    invisible(x)
  }

  png(file = "bkde.png", height = 810, width = 840)
  par(mfrow = c(2, 2))
  plot_bkde2D(dens, col = hsv(seq(from = 2/3, to = 0, length = 33), 0.8, 0.9))
  ## plot_bkde2D(dens, col = rainbow(33, start = 1/3, end = 1, v = 0.8, s = 0.9))
  plot_bkde2D(dens, col = rev(sequential_hcl(33, c = 0, l = c(30, 90), power = c(1/5, 1.5))))
  plot_bkde2D(dens, col = rev(heat.colors(33)))
  plot_bkde2D(dens, col = rev(heat_hcl(33, c = c(80, 30), l = c(30, 90), power = c(1/5, 1.5))))
  dev.off()
}
@
\includegraphics{bkde}
\setkeys{Gin}{width=.47\textwidth}
<<bkde-pal>>=
if(generate_png) {
  png(file = "bkde-pal.png", height = 160, width = 396)
  par(mfrow = c(4, 1), mar = c(0, 0, 0, 0), oma = c(2, 1.5, 0, 1.5))
  plot(0, 0, type="n", xlim = c(0, 0.03), ylim = c(0, 1),
    xaxs = "i", yaxs = "i", axes = FALSE, xlab = "", ylab="")
  abline(v = seq(from = 0, to = 0.03, length.out = 99), lwd = 4, col = 
    hsv(seq(from = 2/3, to = 0, length = 99), 0.8, 0.9))
    ## rainbow(99, start = 1/3, end = 1, v = 0.8, s = 0.9))
  box()
  plot(0, 0, type="n", xlim = c(0, 0.03), ylim = c(0, 1),
    xaxs = "i", yaxs = "i", axes = FALSE, xlab = "", ylab="")
  abline(v = seq(from = 0, to = 0.03, length.out = 99), lwd = 4, col = 
    rev(heat.colors(99)))
  box()
  plot(0, 0, type="n", xlim = c(0, 0.03), ylim = c(0, 1),
    xaxs = "i", yaxs = "i", axes = FALSE, xlab = "", ylab="")
  abline(v = seq(from = 0, to = 0.03, length.out = 99), lwd = 4, col = 
    rev(heat_hcl(99, c = c(80, 30), l = c(30, 90), power = c(1/5, 1.5))))
  box()
  plot(0, 0, type="n", xlim = c(0, 0.03), ylim = c(0, 1),
    xaxs = "i", yaxs = "i", axes = FALSE, xlab = "", ylab="")
  abline(v = seq(from = 0, to = 0.03, length.out = 99), lwd = 4, col = 
    rev(sequential_hcl(99, c = 0, l = c(30, 90), power = c(1/5, 1.5))))
  box()
  axis(1, cex.axis = 1.5)
  dev.off()
}
@
\includegraphics{bkde-pal}
\caption{\label{fig:bkde} Bivariate density estimation for Old Faithful geyser eruptions.}
\end{center}
\end{figure}

A simple and very effective palette for such a display is a set of gray colors
as in the top right panel of Figure~\ref{fig:bkde}. This is often (appropriately) used
in printed papers when the journal does not offer color graphics---however, in journals
that support color graphics, on presentation slides and in interactive usage (in statistical
software packages), many users prefer to have colored displays and most often HSV
palettes (as in the left two panels) are used. These palettes code the variable of interest
by varying hue in an HSV color wheel as done by \citet[Figure~4a]{vcd:Yang+Buckley+Dudoit:2002}
or \citet[Figures~1--4]{vcd:Gneiting+Sevcikova+Percival:2006}. The palette in the upper left panel
codes increasing density by going from a blue to a red hue (via green and yellow)---a similar
strategy are the ``heat colors'' in the lower left panel that increase from yellow to red.
The latter works a bit better than the former, however both palettes exhibit several drawbacks.
The modes in the map look much more like ``rings'' than a smoothly increasing/decreasing 
density. The heatmap looks very flashy which---although good for drawing attention to a plot---makes
it hard to hold the attention of the viewer for a longer time because the large areas shaded
with saturated colors can be distracting and produce after-image effects \citep{vcd:Ihaka:2003}.

In contrast, the gray colors used in the top right panel do not exhibit the same disadvantages,
coding the variable of interest much better and without flashy colors. If, however, the user wants
to add some color to the plot, this could be done by using a better balanced version of the heat
colors (as shown in the bottom right panel). These colors also increase from a yellow to a red
hue while being balanced towards the same gray as in the grayscale palette (i.e., when converted
to a grayscale or printed out on a grayscale printer, the upper and lower right panel would 
essentially look identical). Both palettes have in common that they give increasing perceptual emphasis
to regions with increasing density, resulting in a heatmap that highlights the (small)
interesting high-density regions and not to the large low-density regions surrounding them.

To illustrate sequential palettes,

 Both versions of the heatmap use a sequential
palette as derived in Section~\ref{sec:sequential} balanced towards the same gray levels
with $L \in [30, 90]$ and $p_2 = 1.5$. The sequential palette in the left panel uses only gray
colors (i.e., $C_{\max} = 0$) and the palette in the right panel additionally employs
colors with $H \in [0, 90]$, $C \in [30, 80]$ and $p_1 = 1/5$. The parameters have been 
chosen such that the peaks in the density are brought out very clearly by using a rather
large luminance contrast and $p_2 > 1$.

\setkeys{Gin}{width=\textwidth}
\begin{figure}[t!]
\begin{center}
<<nigeria>>=
if(generate_png) {
  m <- read.bnd("Nigeria/nigeria37.bnd")
  fit <- read.table("Nigeria/reml_purespatial_f_district_spatial.res", header = TRUE)
  fit <- fit[order(fit$district),]

  png(file = "nigeria.png", height = 400, width = 840)
  par(mfrow = c(1, 2), mar = c(2.22, 0.22, 1.02, 0.22))
  plot(m, z = fit$pmode, fill = rev(rainbow(99, start = 0, end = 1/3)),
    zlim = c(-0.65, 0.65), legend = FALSE)
  plot(m, z = fit$pmode, zlim = c(-0.65, 0.65), legend = FALSE,
    fill = rev(diverge_hcl(99, h = c(0, 120), c = 60, l = c(45, 90), power = 1.2)))
  dev.off()
}
@
\includegraphics{nigeria}
\setkeys{Gin}{width=.47\textwidth}
<<nigeria-pal>>=
if(generate_png) {
  png(file = "nigeria-pal.png", height = 80, width = 396)
  par(mfrow = c(2, 1), mar = c(0, 0, 0, 0), oma = c(2, 0.2, 0, 0.2))
  plot(0, 0, type="n", xlim = c(-0.65, 0.65), ylim = c(0, 1),
    xaxs = "i", yaxs = "i", axes = FALSE, xlab = "", ylab="")
  abline(v = seq(from = -0.65, to = 0.65, length.out = 99), lwd = 4, col = 
    rev(rainbow(99, start = 0, end = 1/3)))
  box()
  plot(0, 0, type="n", xlim = c(-0.65, 0.65), ylim = c(0, 1),
    xaxs = "i", yaxs = "i", axes = FALSE, xlab = "", ylab="")
  abline(v = seq(from = -0.65, to = 0.65, length.out = 99), lwd = 4, col = 
    rev(diverge_hcl(99, h = c(0, 120), c = 60, l = c(45, 90), power = 1.2)))
  box()
  axis(1, cex = 1.5)
  dev.off()
}
@
\includegraphics{nigeria-pal}
\caption{\label{fig:nigeria} Posterior mode estimates for childhood mortality in Nigeria.}
\end{center}
\end{figure}

Figure~\ref{fig:nigeria} is taken from \citep[Figure~5, left]{vcd:Kneib:2006}

\setkeys{Gin}{width=\textwidth}
\begin{figure}[p]
\begin{center}
<<seats, fig=TRUE, height=4, width=11>>=
par(mar = rep(0.8, 4), mfrow = c(1, 2))
seats <- structure(c(226, 61, 54, 51, 222),
  .Names = c("CDU/CSU", "FDP",  "Linke", "Gruene", "SPD"))
parties_hcl <- rainbow_hcl(6, c = 60, l = 75)[c(5, 2, 6, 3, 1)]
parties_hsv <- hsv(c(0, 60, 300, 120, 0)/360, v = c(0, 1, 1, 1, 1))
names(parties_hcl) <- names(parties_hsv) <- names(seats)
names(seats)[4] <- "Grüne"
pie(seats, clockwise = TRUE, col = parties_hsv, radius = 1)
pie(seats, clockwise = TRUE, col = parties_hcl, radius = 1)
@
\caption{\label{fig:seats} Seats in the German parliament.}
\end{center}
\end{figure}

\setkeys{Gin}{width=\textwidth}
\begin{figure}[p]
\begin{center}
<<votes, fig=TRUE, height=6, width=14>>=
data("Bundestag2005")
votes <- Bundestag2005[c(1, 3:5, 9, 11, 13:16, 2, 6:8, 10, 12),
                       c("CDU/CSU", "FDP", "SPD", "Gruene", "Linke")]
parties_hcl <- parties_hcl[colnames(votes)]
parties_hsv <- parties_hsv[colnames(votes)]
dimnames(votes)[[2]][c(4, 5)] <- c("Gr", "Li")

grid.newpage()
pushViewport(viewport(layout = grid.layout(1, 2)))
pushViewport(viewport(layout.pos.col=1, layout.pos.row=1))
mosaic(votes, gp = gpar(fill = parties_hsv),
  spacing = spacing_highlighting, labeling = labeling_left,
  labeling_args = list(rot_labels = c(0, 90, 0, 0), pos_labels = "center",
  just_labels = c("center","center","center","right"), varnames = FALSE),
  margins = unit(c(2.5, 1, 1, 12), "lines"),
  keep_aspect_ratio = FALSE, newpage = FALSE)
popViewport()
pushViewport(viewport(layout.pos.col=2, layout.pos.row=1))
mosaic(votes, gp = gpar(fill = parties_hcl),
  spacing = spacing_highlighting, labeling = labeling_left,
  labeling_args = list(rot_labels = c(0, 90, 0, 0), pos_labels = "center",
  just_labels = c("center","center","center","right"), varnames = FALSE),
  margins = unit(c(2.5, 1, 1, 12), "lines"),
  keep_aspect_ratio = FALSE, newpage = FALSE)
popViewport(2)
@
\caption{\label{fig:votes} Votes in the German election 2005 (by province).}
\end{center}
\end{figure}

The first example illustrates qualitative palettes
and visualizes data from the 2005 election for the German parliament ``Bundestag''.
In this election, five parties were able to obtain enough votes to enter the
Bundestag---the distribution of seats is depicted in a pie chart in Figure~\ref{fig:seats}.
The colors used are rough metaphors for the political parties, using a red hue $H = 0$ for the
social democrats SPD, a blue hue $H = 240$ for the conservative CDU/CSU, a yellow hue $H = 60$
for the liberal FDP, a green hue $H = 120$ for the green party ``Die Gr\"unen'' and a purple 
hue $H = 300$ for the leftist party ``Die Linke''. To obtain rather intense colors, we set
chroma to $C = 60$ and luminance to $L = 75$ for all hues
leading to a much better balanced palette compared to the fully saturated
colors usually found in the (German) media (which instead of blue typically use the extreme
``color'' black for the CDU/CSU). The pie chart clearly shows that neither the governing
coalition of SPD and Gr\"une nor the opposition of CDU/CSU and FDP could assemble a majority.
Given that no party would enter a coalition with the leftists, this lead to a big coalition
of CDU/CSU and SPD. Figure~\ref{fig:votes} shows the distribution of votes in this election
stratified by province (Bundesland) in a highlighted mosaic display. The order of provinces is from 
north to south, first for the 10 western provinces (the former Federal Republic of Germany, FRG),
then for the 6 eastern provinces (the former German Democratic Republic, GDR). 
Clearly, the SPD performed better in the north and the CDU/CSU better in the south; furthermore,
Die Linke performed particularly well in the eastern provinces and in Saarland.

\setkeys{Gin}{width=\textwidth}
\begin{figure}[p]
\begin{center}
<<scatter>>=
if(generate_png) {
  set.seed(123)
  vc <- function(var1, var2, cov)
    matrix(c(var1, cov * sqrt(var1 * var2), cov * sqrt(var1 * var2), var2), ncol = 2)
  dat <- rbind(rmvnorm(1000, c(0, 0), vc(0.05, 0.3, -0.95)),
               rmvnorm(1000, c(1.8, 2.5), vc(0.12, 0.25, 0)),
               rmvnorm(1000, c(3.8, 3.5), vc(0.02, 0.15, -0.8)))
  dat <- data.frame(y = dat[,2], x = dat[,1], id = rep(1:3, rep(1000, 3)))

  png(file = "scatter.png", height = 350, width = 840)
  par(mfrow = c(1, 2), mar = rep(0.5, 4))
  plot(y ~ x, data = dat, axes = FALSE, xlab = "", ylab = "", pch = c(3, 22, 1)[id],
    col = hsv(c(1/3, 1/6, 0))[id])
  box()
  plot(y ~ x, data = dat, axes = FALSE, xlab = "", ylab = "", pch = c(3, 22, 1)[id],
    col = hcl2hex(c(120, 60, 0), 70, 65, fix = FALSE)[id])
  box()
  dev.off()
}
@
\includegraphics{scatter}
\caption{\label{fig:scatter} Scatter plot with three clusters.}
\end{center}
\end{figure}

similar to \citet[Figure~3]{vcd:Celeux+Hurn+Robert:2000}

\setkeys{Gin}{width=\textwidth}
\begin{figure}[p]
\begin{center}
<<haireye, fig=TRUE, height=5.5, width=12>>=
he <- margin.table(HairEyeColor, 1:2)
grid.newpage()
pushViewport(viewport(layout = grid.layout(1, 2)))
pushViewport(viewport(layout.pos.col=1, layout.pos.row=1))
mosaic(he, gp = shading_hsv, split_vertical = TRUE, newpage = FALSE, legend = FALSE)
popViewport()
pushViewport(viewport(layout.pos.col=2, layout.pos.row=1))
mosaic(he, gp = shading_hcl, split_vertical = TRUE, newpage = FALSE, legend = FALSE)
popViewport(2)
@
\caption{\label{fig:haireye} Extended mosaic display for haireye data.}
\end{center}
\end{figure}

taken from \citet[Figure~11, left/middle]{vcd:Friendly:2002}

As pointed out in Section~\ref{sec:diverging}, diverging palettes are particularly useful
when visualizing residuals or correlations (with natural neutral value 0) or probabilities 
in 2-class supervised learning (with neutral value 0.5). Examples for both situations are
provided here. Figure~\ref{fig:haireye} visualizes the outcome of a double-blind clinical
trial investigating a new treatment for rheumatoid arthritis. The mosaic rectangles alone
signal that the treatment lead to higher improvement compared to the placebo group; this effect
is shown to be significant by the shading that codes the size of the Pearon residuals.
Positive residuals, corresponding
to more observations in the corresponding cell than expected under independence, are depicted
in blue, negative residuals in red. Light colors signal significance at 10\% level, full colors
significance at 1\% level. Hence, it can be concluded that there are significantly more
marked improvements in the treated group and significantly fewer in the placebo group than
would be expected under independence between treatment and improvement. More details can
be found in \cite{vcd:Zeileis+Meyer+Hornik:2007}.

Both diverging palettes employ a smaller contrast on the luminance axis ($L \in [50, 90]$)
and a larger chroma contrast ($C \in [0, 100]$) compared to the sequential palettes from
the previous example. 
For Figure~\ref{fig:haireye}, the choice of
$p_1$ and $p_2$ is less important because only 5 colors are selected---we use values
$p_1 = p_2 = 1$.

@@

@@

Due to these three color axes, colors are typically described as
locations in 3-dimensional spaces. However, human perception of color
does not correspond to the physiological axes above, but rather to polar
coordinates in the color plane (yellow/blue vs.~green/red) plus a third
light/dark axis. Thus, perceptually-based color spaces try to capture the
perceptual axes
\begin{itemize}
  \item \emph{hue} (dominant wavelength)
  \item \emph{chroma} (colorfulness, intensity of color as compared to gray)
  \item \emph{luminance} (brightness, amount of gray)
\end{itemize}

A popular implementation of such a color space, available in many graphics and
statistics software packages, are HSV colors. They are a simple transformation of RGB colors
and are defined by a triplet $(H, S, V)$ with $H \in [0, 360]$ and $S, V \in [0, 100]$.
%% HSV space has the shape of a single regular cone (often inflated to a regular
%% cylinder). Vertical sections through this space are shown in the upper panel
%% of Figure~\ref{fig:hsv-hcl}, depicting hue and saturation given different value levels.
Although simple to specify and easily available in many computing environments,
HSV colors have a fundamental drawback: its three dimensions map to the three
dimensions of human color perception very poorly. The three dimensions are
confounded:
%% which is most easily seen when converting the vertical sections to 
%% gray scale images in Figure~\ref{fig:hsv-hcl-gray}. Clearly,
the brightness of
colors is not uniform over hues and saturations (given value)---therefore,
HSV colors are often not considered to be perceptually based.

To overcome these drawbacks, various color spaces have been suggested
that properly map to the perception dimensions, the most prominent of which are
the CIELUV and CIELAB spaces developed by the \cite{vcd:CIE:2004}. \cite{vcd:Ihaka:2003}
argues that CIELUV colors are typically preferred for use with emissive technologies
such as computer screens which makes them an obvious candidate for implementation
in statistical software packages. By taking polar coordinates in the UV plane of
CIELUV, HCL colors are obtained, defined by
a triplet $(H, C, L)$ with $H \in [0, 360]$ and $C, L \in [0, 100]$.
%% HCL space has the shape of a distorted double cone: the admissible combinations of
%% chroma and luminance coordinates (within the space's boundaries) depend on the hue chosen.
%% The lower panel of Figure~\ref{fig:hsv-hcl} shows
%% vertical sections through this space: each of the resulting hue/chroma planes (given
%% luminance) is now properly balanced towards the same gray (going from black to white
%% with increasing luminance) which becomes obvious when converting the colors to
%% a gray scale as in Figure~\ref{fig:hsv-hcl-gray}. 
Given a certain luminance $L$, all colors resulting from different combinations
of hue $H$ and chroma $C$ are balanced towards the same gray (and look the same
when converted to a gray scale). However, the admissible combinations of
chroma and luminance coordinates (within the space's boundaries) depend on the hue chosen.
The reason for this is that some hues lead to light and others to dark colors,
e.g., full chroma yellow is brighter (i.e., has higher luminance) than full chroma blue.

The balancing of HCL colors
gives us the opportunity to conveniently choose color palettes which code categorical
and/or numerical information by translating it to paths along the three perceptual axes.
However, some care is required for dealing with the irregular shape of the HCL space
which will be addressed in the following sections.

@@

\setkeys{Gin}{width=\textwidth}
\begin{figure}[p]
\begin{center}
<<wheel, fig=TRUE, height=4, width=11>>=
par(mar = rep(0.8, 4), mfrow = c(1, 2))
par(mfrow = c(1,2), mar = c(1,1,1,1), oma = c(0,0,0,0))
pie(rep(1,6), radius = 1, col = rainbow(6), labels = 360 * 0:5/6)
pie(rep(1,6), radius = 1, col = rainbow_hcl(6, c = 50, l = 70), labels = 360 * 0:5/6)
@
\caption{\label{fig:wheel} HSV-based and HCL-based color wheel.}
\end{center}
\end{figure}

@@

\setkeys{Gin}{width=0.8\textwidth}
\begin{figure}[p]
\begin{center}
<<pal-q-examples, fig=TRUE, height=4, width=6>>=
par(mfrow = c(2, 2), mar = c(0, 0, 3, 2))
pal(rainbow_hcl(4, start = 30, end = 300), main = "dynamic [30, 300]")
pal(rainbow_hcl(4, start = 60, end = 240), main = "harmonic [60, 240]")
pal(rainbow_hcl(4, start = 270, end = 150), main = "cold [270, 150]")
pal(rainbow_hcl(4, start = 90, end = -30), main = "warm [90, -30]")
@
\caption{\label{fig:pal-q-examples} Examples for qualitative palettes. Hue is varied
in different intervals for given $C = 50$ and $L = 70$.}
\end{center}
\end{figure}

@@

\setkeys{Gin}{width=.8\textwidth}
\begin{figure}[p]
\begin{center}
<<pal-s-examples, fig=TRUE, height=5, width=6>>=
par(mfrow = c(5, 1), mar = c(0.25, 0, 0.25, 0))
pal(sequential_hcl(12, c = 0, power = 2.2))
pal(sequential_hcl(12, power = 2.2))
pal(heat_hcl(12, c = c(80, 30), l = c(30, 90), power = c(1/5, 2)))
pal(terrain_hcl(12, c = c(65, 0), l = c(45, 90), power = c(1/2, 1.5)))
pal(rev(heat_hcl(12, h = c(0, -100), l = c(75, 40), c = c(40, 80), power = 1)))
@
\caption{\label{fig:pal-s-examples} Examples for sequential palettes, varying
only luminance (first panel), chroma and luminance (second panel), and hue,
chroma and luminance (remaining panels).}
\end{center}
\end{figure}

@@
